An Implementation of Reference Counting for a Python to x86 Compiler

### 2011-12-15 paper/paper.tex
- Edits through 3.2.4: decrementing reference counts.

Python is a modern, high level language that uses dynamic typing and is run within an interpreter.  It manages memory for the programmer, and is considered by many to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.   Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to deallocate memory.  For language designers, deciding what form of memory management to incorporate into their language can be a difficult task.  The choice will depend on many factors, including the designer's vision on the general usefulness of the language, the environment where the language will be primary used, performance requirements, program semantics, integration with legacy systems and many other factors.  The choice also has significant implications on the implementation and therefore warrants careful consideration.

There are currently two widely used strategies to perform automatic memory management.  One such strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.  This set of objects includes all globals, as well as the objects which are reachable from the current stack frame.  Any objects that are traversed during this operation are considered reachable and therefore cannot be deallocated.  Conversely, any objects not traversed are no longer reachable, and can be deallocated.  The other primary strategy for automated memory management, is reference counting garbage collection. Reference counting works by keeping a count of references to a particular object.  References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs, or when a pointer variable goes out of scope.  When the reference count reaches zero, it is safe to deallocate the object.

The performance characteristics of each strategy can vary considerably, depending on implementation.   Tracing garbage collectors can sometimes be associated with long pause times, where application work must be temporarily stopped to allow the entire contents of the heap to be scanned by the garbage collector.  These long pause times have largely been solved with new algorithms that improve upon the original naive mark-and-sweep algorithm.  It is typical that these algorithms trade-off throughput for responsiveness.  However, these pause times can still be an issue, especially in real time systems, where hard guarantees on latency are needed.  Reference counting has been discussed as a solution to this problem since it incrementally deallocates objects at the program point where all references are destroyed, but classic implementations of reference counting can have serious performance concerns when one deallocation results in cascading deallocations\cite{boehm}.

@@

\section{Related Work}
\label{sec:related}
In general a reference count garbage collector will have shorter pause times for the garbage collection phase, but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.  Reference-counting collectors have been implemented which defer the deallocation, called lazy reference-counting, to some later time and some improvements on such strategies have been made \cite{boehm}.  Without extra work a strictly reference-counting collector will leak memory when references contain cycles, and multiple algorithms have been presented as solutions to this problem. \cite{lin}\cite{trancon}

One potential advantage that reference-counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.  Given the significant cost of garbage collection \cite{hertz} any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.

Real-time systems are rarely implemented in languages which are garbage collected, but reference-counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.  Due to immediate knowledge that an object is no longer in use reference counting garbage collection can allow for destructors or finalizers to be run immediately which can enhance the clean up of system resources.

In \cite{levanoni}, Levanoni, et. al. show that reference counting is a valid approach for multprocessor systems.

\par
\section{Implementation}
\label{sec:implementation}

We start with an existing compiler based upon previous work\cite{siek}.  This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.  Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs, with minimal memory constraints.  After our modifications, the compiler properly maintains reference counts for all objects created, and automatically deallocates objects when there are no longer any references to them.

Implementing reference counting requires a minimum set of steps to be taken.

@@

These steps require modifications to the architecture of our compiler, as well as the supporting runtime library responsible for creating objects and allocating memory.  We discuss these modifications in the following sections.

\subsection{Runtime Modifications}
Our runtime library is implemented in C so that it can be easily called by the assembly programs created by our compiler.  The runtime library contains abstractions such as \afunction{create\_list}, \afunction{create\_dict} and \afunction{create\_object} that take care of allocating and initializing memory for a new object of the specified type.  These functions return a pointer to the memory allocated, and the pointer is \textit{tagged} to indicate that it represents a non-primitive type.  In our reference compiler, any non-primitive object must be a pointer to a C structure called \atype{big\_pyobj}.  To support reference counts, this C structure is modified to include an int field called \afield{ref\_ctr}.  Then all of the functions to create objects are modified to set the initial reference count to zero.

@@

In the \afunction{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.  For the case of a list, we have the \afunction{free\_list} function, which iterates over each element in the list and decrements its reference count.  For a dictionary, we have the \afunction{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.  Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object, if the only reference to it is held by the parent object.  In the worst case, this could lead to a number of deallocations equivalent to the number of nodes in the object graph created by the parent object.

Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such as reference is established.  For example, an assignment to a list or dictionary subscript results in a call to the \afunction{set\_subscript} runtime function.  This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and then increment the reference count for the new object that is assigned to the given subscript.  Similarly, a closure maintains a reference to its list of free variables, and a class object maintains a reference to its base classes.  In each case, we modify the appropriate runtime function to ensure that these reference counts are correctly maintained.

In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).  For this reason, significant changes were made to the runtime to ensure that any type that references another type, has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.  To ensure that we did not make any mistakes in this process, we create a runtime test suite that uses the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.  This runtime test suite is important to unit test the runtime library, before attempting to use it from our compiler.

Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.  These abstractions are part of our runtime and provide us the following benefits.

@@

The abstractions are implemented in a new module appropriately called \texttt{pymem}.  Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}. As a result of this, we can determine exactly how much memory has been allocated and what objects have still not been deallocated, by calling the \afunction{pymem\_print\_statistics} function at any given program point.  In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program. } to ensure that all memory has been deallocated.  Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.

The \afunction{pymem\_new} routine works by allocating slightly more memory than is requested by the user.  This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \afunction{malloc} and the time of allocation.  The caller is then returned the portion of memory immediately after this pointer.  Subsequent calls to \afunction{pymem\_free} then have access to the linked list node via the pointer, and can update the node to mark the memory as deallocated, as well as the deallocation time.  The memory can then be freed independently of the linked list containing the statistics.

@@

Figure~\ref{fig-comparch} shows the compiler data flow after modifying our compiler for reference counting.  Each stage operates on an abstract syntax tree (AST).  The relevant new stages are the intermediate flatten stage, and the reference counting AST transformation stage.

@@

Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it, and decrement the reference count when a reference is destroyed.  References are created by assigning a memory location (which contains an object) to a variable.  References are destroyed when a variable is changed to refer to a different object, or when a variable ceases to exist when a program reaches the end of a given scope.

The most natural way to implement this in our compiler, is to transform our input programs AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.  This is the approach we took, but there are some subtleties with respect to where this operation is placed within the compiler.  Our initial hypothesis was that we could easily add the increment and decrement reference counter operations in the existing ``Flatten AST'' stage.  We quickly realized that because this stage is after type explication, there exist variables which are not \textit{tagged} with a type.  Since we only want to decrement reference counts for \textit{tagged} types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types. However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \atype{int} and \atype{boolean}.}, the only option is to put this transformation stage before type explication.  In addition to modifying the AST prior before/after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.

\subsubsection{Expressions}
Prior to our explicate phase, complex expressions are not flattened.  This leads to problems where memory is leaked, since no assignments ever occur to temporary objects.  Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between to lists to the variable \avariable{x}.  The assignment to the variable \avariable{x} results in a corresponding increment reference count on the result of the list addition.  However, the temporary lists, \aliteral{[1]} and \aliteral{[2]} are leaked if we don't take any additional steps.  To overcome this problem, we introduce a intermediate flattening stage, prior to our reference counting AST transformation.   This results in the corresponding code given in Listing~\ref{lst:flattened} after flattening.  These temporary objects are then deallocated when their corresponding references go out of scope.

@@

\subsubsection{Temporary variables}
As a result of introducing the intermediate flattening stage, we introduce additional references to an object which hang around until the end of the given scope.  This is undesirable because it changes the deallocation behavior of the program from the expected behavior.  For original input program in listing~\ref{lst:simpleleak}, the expectation is that when the variable \avariable{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.  However, in the flattened program, this is no longer the case since the \avariable{tmp2} variable continues to reference the object.  The deallocation is therefore delayed until the end of the current scope.  To handle this situation, we keep track of all temporary variables introduced as a result of flattening an expression.  These temporary variables are then assigned a value of zero immediately after the statement containing the expression in which they were introduced.  This will lead to an immediate decrement of the reference count after the reference counting AST transformation.  Listing~\ref{lst:fixflattened} shows the result of the flattening transformation after this change is made.

>>>>

Python is a modern, high level language that uses dynamic typing and is run within an interpreter.  It manages memory for the programmer and is considered to be a good ``prototyping'' language, in which programmers can quickly and easily develop an application.   Automatic memory management plays a large part in supporting this role, as it relieves the programmer from the tedious task of remembering when to deallocate memory.  For language designers, deciding what form of memory management to incorporate into their language can be a difficult task.  The choice will depend on many factors, including the designer's vision on the general usefulness of the language, the environment where the language will be primary used, performance requirements, program semantics, integration with legacy systems and many other factors.  The choice also has significant implications on the implementation and therefore warrants careful consideration.

There are currently two widely used strategies to perform automatic memory management.  One strategy is tracing garbage collection, which periodically traverses the in-memory object graph starting from a currently reachable set of ``root'' objects.  This set of objects includes all global objects, as well as the objects which are reachable from the current stack frame.  Any objects that are traversed during this operation are marked as reachable and therefore cannot be deallocated.  Any object which is not marked is a candidate for deallocation.  The other primary strategy for automated memory management is reference counting garbage collection. Reference counting works by keeping a count of references to a particular object.  References are tracked by incrementing and decrementing an object's reference count when an assignment to a pointer occurs or when a pointer variable goes out of scope.  When the reference count reaches zero, it is safe to deallocate the object.

The performance characteristics of each strategy can vary considerably depending on implementation.   Tracing garbage collectors can sometimes be associated with long pause times, where application work must be temporarily stopped to allow the entire contents of the heap to be scanned by the garbage collector.  These long pause times have largely been solved with new algorithms that improve upon the original naive mark-and-sweep algorithm.  It is typical that these algorithms trade-off throughput for responsiveness.  However, these pause times can still be an issue, especially in real time systems, where hard guarantees on latency are needed.  Reference counting has been discussed as a solution to this problem since it incrementally deallocates objects at the program point where all references are destroyed, but classic implementations of reference counting can have serious performance concerns when one deallocation results in cascading deallocations\cite{boehm}.

@@

\section{Related Work}
\label{sec:related}
In general, a reference count garbage collector will have shorter pause times for the garbage collection phase but will incur a higher performance penalty \cite{joisha}\cite{blackburn}.  To address these performance concerns, lazy reference counting collectors have been implemented which defer deallocation to some later time.  Improvements on such strategies are proposed by Boehm \cite{boehm}. In \cite{levanoni}, Levanoni, et. al. show that reference counting is a valid approach for multprocessor systems.

One potential advantage that reference counting garbage collectors have is the ability to be implemented in hardware \cite{joao}.  Given the significant cost of garbage collection \cite{hertz}, any hardware assisted acceleration would be highly desirable if it is flexible enough to adapt across multiple garbage collection techniques.  Real-time systems are often very concerned with latency, and therefore do not typically make use of garbage collection.  However, reference counting garbage collection has been adapted to hard real-time systems \cite{ritzaou}.

Because reference counting garbage collection can deallocate an object at the specific program point where there are no longer any references, it can allow for finalizers to be run immediately.  This makes the use of finalization a valid approach to releasing system resources.

One of the primary disadvantages of a strictly reference counting collector is that it will leak memory when references contain cycles.  Multiple algorithms have been presented as solutions to this problem \cite{lin}\cite{trancon}.

\section{Implementation}
\label{sec:implementation}

We start with an existing compiler based upon previous work \cite{siek}.  This compiler converts a subset of the Python language to x86 assembly and supports one of the core features of Python, namely that everything is a first class object.  Prior to our modifications, the compiler does not perform any memory management, resulting in a language that is only useful for short lived programs with minimal memory constraints.  After our modifications, the compiler properly maintains reference counts for all objects created and automatically deallocates objects when there are no longer any references to them.

Implementing reference counting requires the following set of steps to be taken:

@@

These steps require modifications to the architecture of our compiler and to the supporting runtime library responsible for creating objects and allocating memory.  We will discuss these modifications in the following sections.

\subsection{Runtime Modifications}
Our runtime library is implemented in C so that it can be easily called by the assembly programs created by our compiler.  The runtime library contains abstractions such as \afunction{create\_list}, \afunction{create\_dict} and \afunction{create\_object} that take care of allocating and initializing memory for a new object of the specified type.  These functions return a pointer to the memory allocated, and the pointer is \textit{tagged} to indicate that it represents a non-primitive type.  In our reference compiler, any non-primitive object must be a pointer to a C structure called \atype{big\_pyobj}.  To support reference counts, this C structure is modified to include an int field called \afield{ref\_ctr}.  Then, all of the functions to create objects are modified to set the initial reference count to zero.

@@

In the \afunction{free\_\textit{type}} functions, it is necessary to add further calls to decrement the reference counter of any objects referenced by the object being deallocated.  In the case of a list, we have the \afunction{free\_list} function, which iterates over each element in the list and decrements its reference count.  For a dictionary, we have the \afunction{free\_dict} function, which decrements the reference counter on keys as well as values contained in the dictionary, since both can be objects.  Note that for any case of an object that references other objects, this can can result in deallocation of the referenced object if the only reference to it is held by the parent object.  In the worst case, this could lead to a number of deallocations equivalent to the number of nodes in the object graph created by the parent object.

Since every object in our system may contain references to another object, we modify our runtime to correctly increment the reference count when such a reference is established.  For example, an assignment to a list or dictionary subscript results in a call to the \afunction{set\_subscript} runtime function.  This function is modified to first decrement the reference count for any object that may have been previously referenced by the given subscript, and to then increment the reference count for the new object that is assigned to the given subscript.  Similarly, a closure maintains a reference to its list of free variables, and a class object maintains a reference to its base classes.  In each case, we modify the appropriate runtime function to ensure that these reference counts are correctly maintained.

In some cases, the type definition for an object in our runtime did not aggregate a \atype{big\_pyobj} type, and instead maintained only a pointer to the non-generic version of the type (e.g., \atype{class\_struct}).  For this reason, significant changes were made to the runtime to ensure that any type that references another type has a generic pointer to a \atype{big\_pyobj} structure, so that the reference counts can be correctly maintained.  To ensure that we did not make any mistakes in this process, we created a runtime test suite that used the abstractions provided by the runtime library to create objects, increment and decrement their reference counts and ensure that all objects are deallocated when their reference count reaches zero.  This runtime test suite is important to unit test the runtime library before attempting to use it from our compiler.

Finally, in order to verify the correctness of our implementation, we create a set of abstractions on top of the system allocation functions \afunction{malloc} and \afunction{free}.  These abstractions are part of our runtime and provide us the following benefits:

@@

The abstractions are implemented in a new module appropriately called \texttt{pymem}.  Every allocation of an object in our runtime is modified to use the \afunction{pymem\_new} and \afunction{pymem\_free} instead of \afunction{malloc} and \afunction{free}. As a result, we can determine exactly how much memory has been allocated and what objects have still not been deallocated by calling the \afunction{pymem\_print\_statistics} function at any given program point.  In practice, this is only called at the end of each program\footnote{This is achieved by modifying the x86 generation phase of our compiler to include calls to set up our \texttt{pymem} module and print statistics at the end of the program. } to ensure that all memory has been deallocated.  Since we can track the time of allocation and deallocation, this module can be beneficial for investigating object lifetime as well.

The \afunction{pymem\_new} routine works by allocating slightly more memory than is requested by the user.  This extra memory is used to store a pointer to a linked list node that contains the details of the allocation, including the type of object being created, the requested size, the pointer returned by \afunction{malloc} and the time of allocation.  The caller is then returned the portion of memory immediately after this pointer.  Subsequent calls to \afunction{pymem\_free} then have access to the statistics in the linked list node via the pointer.  These calls will then update the node to mark the memory as deallocated and record the deallocation time.  The memory can then be freed independently of the linked list containing the statistics.

@@

Figure~\ref{fig-comparch} shows the compiler data flow after modifying our compiler for reference counting.  Each stage operates on an abstract syntax tree (AST).  The relevant new stages are the intermediate flatten stage and the reference counting AST transformation stage.

@@

Recall that in order to properly maintain reference counts, we need to increment the reference count for an object when a reference is created for it and decrement the reference count when a reference is destroyed.  References are created by assigning a memory location (which contains an object) to a variable.  References are destroyed when a variable is changed to refer to a different object, or when a variable ceases to exist when a program reaches the end of a given scope.

The most natural way to implement this in our compiler is to transform our input program's AST such that any assignment to a variable is preceded by a decrement to the reference count for the object the variable refers to, and is succeeded by an increment to the reference count for the newly assigned object.  This is the approach we took, but there are some subtleties with respect to where this operation is placed within the compiler.  Our initial hypothesis was that we could easily add the increment and decrement reference counter operations in the existing ``Flatten AST'' stage.  We quickly realized that because this stage is after type explication, there exist variables which are not tagged with a type.  Since we only want to decrement reference counts for tagged types\footnote{Actually, we only want to modify reference counts for a subset of tagged types that only includes ``big'' types. However, to simplify our implementation, we transform our AST to always call our increment and decrement runtime functions, but these functions will ignore the primitive types of \atype{int} and \atype{boolean}.}, the only option is to put this transformation stage before type explication.  In addition to modifying the AST before and after an assignment, the reference counting transformation is also responsible for determining which variables are local to a given scope and decrementing the counts for these references at the end of the given program scope.

\subsubsection{Expressions}
Prior to our explicate phase, complex expressions are not flattened.  This leads to problems where memory is leaked, since no assignments are ever made to temporary objects.  Listing~\ref{lst:simpleleak} shows a program that assigns the result of addition between two lists to the variable \avariable{x}.  The assignment to the variable \avariable{x} results in a corresponding increment reference count on the result of the list addition.  However, the temporary lists, \aliteral{[1]} and \aliteral{[2]} are leaked if we don't take any additional steps.  To overcome this problem, we introduce an intermediate flattening stage prior to our reference counting AST transformation.   This results in the corresponding code given in Listing~\ref{lst:flattened} after flattening.  These temporary objects are then deallocated when their corresponding references go out of scope.

@@

\subsubsection{Temporary variables}
As a result of introducing the intermediate flattening stage, we introduce additional references to an object which hang around until the end of the given scope.  This is undesirable because it changes the deallocation behavior of the program from the expected behavior.  For the original input program in listing~\ref{lst:simpleleak}, the expectation is that when the variable \avariable{x} is assigned the value zero, the reference count for the list object should be decremented to zero and the object deallocated.  However, in the flattened program, this is no longer the case since the \avariable{tmp2} variable continues to reference the object.  The deallocation is therefore delayed until the end of the current scope.  To handle this situation, we keep track of all temporary variables introduced as a result of flattening an expression.  These temporary variables are then assigned a value of zero immediately after the statement containing the expression in which they were introduced.  This will lead to an immediate decrement of the reference count after the reference counting AST transformation.  Listing~\ref{lst:fixflattened} shows the result of the flattening transformation after this change is made.
